On multilevel constraint decomposition methods \\ for nonlinear variational inequalities

### 2022-07-17 paper/mcd2.tex
by splitting algs and commenting, I am really clear on them

Note that $\cK_i \not\subset \cK$ in most cases.  In Example \ref{ex:domaindecomposition}, if $\psi$ is positive over portions of $\Omega$ where the decomposition into overlapping subdomains $\Omega_i$ is nontrivial, then $\cK_i \not\subset \cK$, and similarly for Example \ref{ex:frequencydecomposition}.  Instead, the important inclusion is $\cK_i \subset \cV_i$.

Algorithm \ref{alg:basiccd} below states the basic CD algorithm as an iteration which solves smaller VI problems over each set $\cK_i$.  (By contrast, the algorithms in \cite{Tai2003} solve optimization problems over each $\cK_i$.)  The algorithm starts from a current iterate $u \in \cK$ and computes a new iterate $w\in\cK$, an improved approximation of the solution $u^* \in \cK$ of \eqref{eq:vi}, using a damping parameter $0<\alpha\le 1$.  There are parallel (additive) and successive (multiplicative) versions of the algorithm, generalizing the Jacobi and Gauss-Seidel iterations \cite{Greenbaum1997}, respectively.  For the parallel version the \textbf{for} loop can be computed in any order.

\begin{pseudofloat}[H]
\begin{pseudo*}
\pr{cditeration}(u\in\cK)\text{:} \\+
    if \pr{parallel}: \\+
        for $i \in \{0,\dots,m-1\}$: \\+
            $\hat w_i\in \cK_i$: \\+
                 $\boxed{\ip{f(u - R_i u + \hat w_i)}{v_i-\hat w_i} \ge g(v_i-\hat w_i)} \quad \forall v_i\in \cK_i$ \\---
    else: \\+
        for $i = 0,\dots,m-1$: \\+
            $\hat w_i\in \cK_i$: \\+
                $\displaystyle \boxed{\ip{f\Big(\sum_{j<i} w_j + \hat w_i + \sum_{j>i} R_j u\Big)}{v_i-\hat w_i} \ge g(v_i-\hat w_i)} \quad \forall v_i\in \cK_i$ \\-
            $w_i = (1-\alpha) R_i u + \alpha \hat w_i\in\cK_i$ \\--
    $\hat w = \sum_i \hat w_i\in\cK$ \\
    return $w=(1-\alpha) u + \alpha \hat w\in\cK$
\end{pseudo*}
\caption{The basic constraint decomposition (CD) algorithm for VI problem \eqref{eq:vi}.}
\label{alg:basiccd}
\end{pseudofloat}

The reader may confirm that, inside each boxed VI in Algorithm \ref{alg:basiccd}, the argument of $f$ is an element of $\cK$.  While the difference $v_i - \hat w_i \in \cV_i$ appears, note that by \eqref{eq:constraintdecomp} and \eqref{eq:constraintrestrictionsum} one may write it as a difference of admissible vectors (i.e.~from $\cK$), namely
\begin{align*}
[u - R_i u + v_i] - [u - R_i u + \hat w_i] &= v_i - \hat w_i, \label{eq:admissibledifference} \\
\left[\sum_{j<i} w_j + v_i + \sum_{j>i} R_j u\right] - \left[\sum_{j<i} w_j + \hat w_i + \sum_{j>i} R_j u\right] &= v_i - \hat w_i,  \notag
\end{align*}
for the two versions, respectively.

In other words, we solve VI \eqref{eq:vi} over perturbations from the sets $\cK_i$.  Define
\begin{equation}
e_i = \hat w_i - R_i u \in \cV_i \label{eq:ithupdate}
\end{equation}
as the $i$th update from a pass through either \textbf{for} loop in Algorithm \ref{alg:basiccd}.  Then $\hat w = u + \sum_{i} e_i$ and $w = u + \alpha \sum_i e_i$.  Easy identities $u - R_i u + \hat w_i = u + e_i$ and $\hat w = u^* + \sum_i \hat w_i - R_i u^*$ also hold.  We will use these identities in Section \ref{sec:convergence}.

The convergence results in Section \ref{sec:convergence} require substantial damping when applying the parallel algorithm, i.e.~$\alpha \le 1/m$, the same as in \cite{Tai2003}, but the successive version can be shown to converge without damping ($\alpha=1$).  In Section \ref{sec:results} we demonstrate practical convergence for a larger range of $\alpha$ than suggested by the theory.  % HOPE

In order for Algorithm \ref{alg:basiccd} to have a practical and efficient FE implementation (Section \ref{sec:multilevel}), we need additional notions when $f$ has any of the following properties: \emph{(i)} $f$ is defined only on $\cK$, or \emph{(ii)} $f$ is nonlinear, or \emph{(iii)} $f$ is non-local.  The issue of practical implementation for any such $f$ seems not to have been addressed in the CD literature, which is why we call Algorithm \ref{alg:basiccd} the ``basic'' algorithm.  In particular  \cite{GraeserKornhuber2009,Tai2003} only apply the CD method to the classical obstacle problem.  The following example explains the simplifications available in that easy case.

\begin{example}  \label{ex:fnice} Suppose $f:\cV \to \cV'$ is linear and defined on all of $\cV$.  Furthermore suppose $f$ is local in the sense that a basis $\{\phi_i\}$ of $\cV_i$ exists, with each support small in $\Omega$, such that $\ip{f(\phi_i)}{z}$ can be computed by an integral over the support of $\phi_i$.  (When $\cV$ is an FE space then this is the usual case for evaluating PDE weak forms over a basis of hat functions; see Section \ref{sec:multilevel}.  Constrast integral operators where $\ip{f(\phi_i)}{z}$ requires an integral over $\Omega$ even if the support of $\phi_i$ is small.)  Considering only the parallel version of Algorithm \ref{alg:basiccd} for simplicity, the boxed VI over $\cK_i$ can be written as
\begin{equation}
\ip{f(e_i)}{v_i-\hat w_i} \ge \tilde g(v_i-\hat w_i) \label{eq:linearlocalvi}
\end{equation}
where $\tilde g(z) = g(z) - \ip{f(u)}{z}$.  Noting $e_i = \hat w_i - R_i u \in \cV_i \notin \cK$, in general, VIs \eqref{eq:linearlocalvi} will only make sense because $f$ is defined over $\cV$.  Each problem \eqref{eq:linearlocalvi} can be solved using a stored residual $\ip{f(u)}{\cdot}$ (already included into the source term).  The VI \eqref{eq:vi} is then approximately solved in any incremental and efficient manner, by computations over the basis supports, during one application of \pr{cditeration}.
\end{example}

In other words, when $f$ has all the nice properties proposed in Example \ref{ex:fnice} then the decomposed VI problems can be computed from inexpensive data.  Note that the actual solution method for VI \eqref{eq:linearlocalvi} is not the concern here; we are observing that the data of problem \eqref{eq:linearlocalvi}, and the cost of residual evaluation, is small in practice.  A solver which would work for Example \ref{ex:fnice} can thus have implementation efficiencies which are unavailable in general.

Additional ideas, like those in Section \ref{sec:multilevel}, are needed for practical application of CD methods to more difficult problems.  In particular, we will extend the basic Algorithm \ref{alg:basiccd} to nonlinear $f$ by applying the full approximation storage (FAS) idea of Brandt \cite{Brandt1977}, in which case the decomposed problems only evaluate $f$ over (admissible) elements of $\mathcal{K}$, and we propose restrictions which allow inexpensive approximations of the decomposed problems.  However, extending the algorithm to non-local residual functionals is a topic for future research.

>>>>

Note that $\cK_i \not\subset \cK$ in many cases.  Specifically in Example \ref{ex:domaindecomposition}, if $\psi$ is positive over portions of $\Omega$ where the decomposition into overlapping subdomains $\Omega_i$ is nontrivial, then $\cK_i \not\subset \cK$, and similarly for Example \ref{ex:frequencydecomposition}.  The important inclusion is $\cK_i \subset \cV_i$.

Algorithms \ref{alg:basiccd-add} and \ref{alg:basiccd-mult} below state the basic CD method as iterations which improve a current iterate $u \in \cK$ by solving ``smaller'' VI problems over each subset $\cK_i$.  The output $w\in\cK$ should be closer to the solution $u^* \in \cK$ of \eqref{eq:vi}.

\begin{pseudofloat}[H]
\begin{pseudo*}
\pr{cd-add}(u\in\cK)\text{:} \\+
    for $i \in \{0,\dots,m-1\}$: \\+
        \rm{find} $\hat w_i\in \cK_i$ \rm{so that for all} $v_i\in \cK_i$ \rm{we have} \\+
            $\boxed{{\large \strut} \ip{f(u - R_i u + \hat w_i)}{v_i-\hat w_i} \ge g(v_i-\hat w_i)}$ \\--
    $\hat w = \sum_i \hat w_i\in\cK$ \\
    return $w=(1-\alpha) u + \alpha \hat w\in\cK$
\end{pseudo*}
\caption{Additive constraint decomposition for VI problem \eqref{eq:vi}.}
\label{alg:basiccd-add}
\end{pseudofloat}

The \textbf{for} loop in Algorithm \ref{alg:basiccd-add}, the additive or parallel CD method, can be computed in any order.  Note that $u-R_iu+\hat w_i$ involves removing the part of $u$ which lies in $\mathcal{K}_i$ before adding-back the improved subset solution $\hat w_i$.

The following multiplicative (successive) version orders the sets $\mathcal{K}_i$, and each subset solution updates the global iterate.  While Algorithm \ref{alg:basiccd-add} generalizes a Jacobi iteration, this one is like Gauss-Seidel.  The sum $\sum_{j>i} R_j u$ should be read as ``keep the parts of the current iterate which we have not yet improved.''

\begin{pseudofloat}[H]
\begin{pseudo*}
\pr{cd-mult}(u\in\cK)\text{:} \\+
    for $i = 0,\dots,m-1$: \\+
        \rm{find} $\hat w_i\in \cK_i$ \rm{so that for all} $v_i\in \cK_i$ \rm{we have} \\+
            $\displaystyle \boxed{\ip{f\Big(\sum_{j<i} w_j + \hat w_i + \sum_{j>i} R_j u\Big)}{v_i-\hat w_i} \ge g(v_i-\hat w_i)}$ \\-
            $w_i = (1-\alpha) R_i u + \alpha \hat w_i\in\cK_i$ \\-
    $\hat w = \sum_i \hat w_i\in\cK$ \\
    return $w=(1-\alpha) u + \alpha \hat w\in\cK$
\end{pseudo*}
\caption{Multiplicative constraint decomposition for VI problem \eqref{eq:vi}.}
\label{alg:basiccd-mult}
\end{pseudofloat}

Note the use of a damping parameter $0<\alpha\le 1$.  In Algorithm \ref{alg:basiccd-add} its role is obvious, as we keep part of the current iterate if $\alpha<1$, but in Algorithm \ref{alg:basiccd-mult} we only update the iterate ``so far'', namely $w_i$, using the same damping that will be applied in generating the final output $w$.

In each boxed VI above the argument of $f$ is an element of $\cK$, as the reader should confirm.  Likewise, by \eqref{eq:constraintdecomp} and \eqref{eq:constraintrestrictionsum} one may write the test element $v_i - \hat w_i \in \cV_i$ as a difference of admissible vectors, from $\cK$, namely
\begin{align*}
[u - R_i u + v_i] - [u - R_i u + \hat w_i] &= v_i - \hat w_i, \label{eq:admissibledifference} \\
\left[\sum_{j<i} w_j + v_i + \sum_{j>i} R_j u\right] - \left[\sum_{j<i} w_j + \hat w_i + \sum_{j>i} R_j u\right] &= v_i - \hat w_i,  \notag
\end{align*}
for the two Algorithms respectively.

Define
\begin{equation}
e_i = \hat w_i - R_i u \in \cV_i \label{eq:ithupdate}
\end{equation}
as the $i$th subset update in either Algorithm.  Then, as the reader may check, $\hat w = u + \sum_{i} e_i$ and $w = u + \alpha \sum_i e_i$.  The easy identity $\hat w = u^* + \sum_i \hat w_i - R_i u^*$ also holds.  We will use these identities in Section \ref{sec:convergence}.

% FOLLOWING CONTAINS ASPIRATIONS
The convergence results in Section \ref{sec:convergence} require substantial damping when applying additive Algorithm \ref{alg:basiccd-add}, namely we require $\alpha \le 1/m$ as does \cite{Tai2003}.  Algorithm \ref{alg:basiccd-mult} can be shown to converge without damping ($\alpha=1$).  In Section \ref{sec:results} we demonstrate practical convergence for larger ranges of $\alpha$ than suggested by the theory.

We refer to Algorithms \ref{alg:basiccd-add} and \ref{alg:basiccd-mult} as the \emph{basic CD method}.  To construct efficient FE implementations (Section \ref{sec:multilevel}) we will need additional notions.  While the above Algorithms are indeed meaningful even when $f$ is nonlinear, non-local, or defined only on $\cK$, practical implementation for such general problems seems not to have been addressed in the literature.  In particular, references \cite{GraeserKornhuber2009,Tai2003} only apply the CD method to the classical obstacle problem.  The following example explains the implementation-relevant simplifications available in that case.

\begin{example}  \label{ex:fnice} Suppose $f:\cV \to \cV'$ is linear and defined on all of $\cV$.  Furthermore suppose $f$ is local in the sense that a basis $\{\phi_i\}$ of $\cV_i$ exists, with $\supp \phi_i$ compactly-contained in $\Omega$, such that for any $z\in\mathcal{V}$ the value $\ip{f(\phi_i)}{z}$ can be computed by an integral over the support of $\phi_i$.  Considering only additive Algorithm \ref{alg:basiccd-add} for simplicity, by linearity the boxed VI over $\cK_i$ can be written as
\begin{equation}
\ip{f(e_i)}{v_i-\hat w_i} \ge \tilde g(v_i-\hat w_i) \label{eq:linearlocalvi}
\end{equation}
where $\tilde g(z) = g(z) - \ip{f(u)}{z}$.  Note $e_i = \hat w_i - R_i u \in \cV_i \notin \cK$, in general, but VIs \eqref{eq:linearlocalvi} make sense because $f$ is defined over $\cV$.  Each problem \eqref{eq:linearlocalvi} can be solved using a stored residual $\ip{f(u)}{\cdot}$, namely as included here into the source term.  The VI \eqref{eq:vi} is then approximately solved in any incremental and efficient manner, by computations over the basis supports.
\end{example}

The assumption that $\ip{f(\phi_i)}{z}$ can be evaluated over the support of $\phi_i$ is satisfied when $f$ is a linear differential operator, $\cV$ is an FE space, and $\phi_i$ are hat functions; see Section \ref{sec:multilevel}.  Constrast non-local integral operators where $\ip{f(\phi_i)}{z}$ requires an integral over $\Omega$ even if $\phi_i$ is is a hat function with small support.

When $f$ has all the nice properties listed in Example \ref{ex:fnice} then a solver can take advantage of implementation efficiencies which are unavailable in general.  The actual solution process for VI \eqref{eq:linearlocalvi} is not the concern here; we are observing that the data of problem \eqref{eq:linearlocalvi}, and the cost of residual evaluation, are smaller than in the general case.  For more difficult problems additional ideas, illustrated in Section \ref{sec:multilevel}, are needed for practical implementation.  We will extend the basic method to nonlinear $f$ by applying the full approximation storage (FAS) idea of Brandt \cite{Brandt1977}, and the decomposed problems will only evaluate $f$ over (admissible) elements of $\mathcal{K}$.  However, extending the algorithm to non-local functionals $f$ is a topic for future research.
